---
title: "OOPSLA20 Artifact - Designing Types for R, Empirically"
authors: "Alexi Turcotte, Aviral Goel, Filip Krikava and Jan Vitek"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Artifact Description

This is the artifact for the OOPSLA 2020 paper *Designing Types for R, Empirically*.
The aim is to:

1.  show the tools that were developed to infer and assert types for R
    functions,
2.  demonstrate the entire pipeline used to infer types for R packages.

The artifact is composed of two parts: 

1. a getting started guide that contains the setup instructions and a small
   experiment to verify that all works, and
2. a step-by-step instructions on how the tooling developed for this paper work
   and a complete pipeline which reproduces the data reported in the paper.

## Requirements 

The pipeline depends on a number of tools and R packages:

- git >= 2
- bash >= 4
- GNU parallel >= 20190322
- GNU bison >= 3.5
- R == 3.5.0
- R packages
  - [rapr](https://github.com/PRL-PRG/rapr)_for running the experiments
  - [contractr](https://github.com/PRL-PRG/contractr)
  - ...


To make it more convenient we have build a docker image that has all these dependencies installed.
The image is available on [Docker HUB](https://hub.docker.com/r/prlprg/oopsla20-typer).
This image can be either pulled directly from the Hub or <a href="#building-image-locally">built locally</a>.
To run the image, you will need [docker community edition](https://docs.docker.com/install/) version 18+ and bash.


This artifact requires about ~13GB of free space.


## Getting Started Guide

Open `README.Rmd` by either navigating to `File` -> `Open File...` menu item or by running:

```{r eval=FALSE}
rstudioapi::navigateToFile("~/README.Rmd")
```

- Getting the docker image
- Connecting to R studio
- Running the pipeline on a few packages
  - a set of 10 packages that runs quick?

Check that Rstudio can see the installed corpus.
The number of installed packages should be 603.

```{r}
length(installed.packages()[,1])
```

This concludes the getting started guide.

## Step-by-step Instructions

### Quick Tutorial

#### Types for R functions

#### Types for R packages

#### Inferring Types from R Code

### Redoing the Paper Experiment

#### Corpus

```{r}
rmarkdown::render("~/typeR/analysis/corpus-analysis.Rmd")
rstudioapi::viewer("~/typeR/analysis/corpus-analysis.html")
```

```{r eval=FALSE}
rstudioapi::navigateToFile("~/typeR/analysis/corpus-analysis.Rmd")
```

## Artifact Structure

TODO

## Building Image Locally
<a name="#building-image-locally"/>

To build image locally you can either use make by running:

```{sh eval=FALSE}
make .docker-build
```

## Toubleshooting

### Docker on OSX

It is better to use homebrew cask to install docker:

```sh
brew cask install docker
```

in case you see `docker: Cannot connect to the Docker daemon at unix:///var/run/docker.sock. Is the docker daemon running?.` error message
cf: https://stackoverflow.com/a/44719239

### Docker on Linux

In some distribution the package does not add the current user to `docker` group.
In this case, either add yourself to `docker` group or run all docker-related command with `sudo`.
