---
title: "Designing Types for R, Empirically -- Paper Reproduction"
output: html_notebook
params: 
  data_dir: ~/typeR/type-inference/data-corpus
  script_dir: ~/typeR/type-inference/scripts
---

This file contains the data and code for producing select Figures and Tables from the paper.
As the data is available, feel free to play with the data frames and explore!
Configure the parameter `paper_eval` at the top of the file to be `FALSE` if you'd like to test the fresh data you ran.

This first bit of code is to load all the data and dependencies.

```{r}
# Load the initial data.
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(purrr))
library(readr)
library(runr)
library(tidyr)

# Data manipulation functions.
source(file.path(params$script_dir, "data-analysis-lib.R"))

# Other functions for data processing.
simplify_types <- function(type) {
    if(type %in% c("any", "externalptr", "expression", "pairlist")) {
        "any"
    }
    else {
        type
    }
}

infer_kind <- function(type) {
    if(type %in% c("integer", "double", "character", "logical", "complex", "raw")) {
        "scalar"
    }
    else if(type %in% c("integer[]", "double[]", "character[]", "logical[]", "complex[]", "raw[]")) {
        "vector"
    }
    else if(type %in% c("^integer[]", "^double[]", "^character[]", "^logical[]", "^complex[]", 
                        "^integer", "^double", "^character", "^logical", "^complex")) {
        "^vector"
    }
    else if(str_detect(type, "^list")) {
        "list"
    }
    else if(str_detect(type, "^struct")) {
        "struct"
    }
    else if(str_detect(type, "^class")) {
        "class"
    }
    else if(type == "any") {
        "any"
    }
    else if(type == "null") {
        "null"
    }
    else if(type == "...") {
        "..."
    }
    else if(type == "environment") {
        "environment"
    }
    else {
        "other"
    }
}

# Data with types of function arguments and returns.
type_analysis_data <- read_csv(file.path(params$data_dir, "type-traces.csv"))

function_arg_counts <- type_analysis_data %>% group_by(package_name, function_name, parameter_position) %>% summarize %>% ungroup

# Include type "kinds"
type_analysis_data <-
    type_analysis_data %>%
    mutate(type = map_chr(type, simplify_types)) %>%
    mutate(kind = map_chr(type, infer_kind))

# Find incompatible functions.
type_compatibility <-
    type_analysis_data %>%
    group_by(package_name, function_name) %>%
    summarize(compatible = length(unique(outer_alternative)) == 1) %>%
    ungroup()

incompatible_functions <-
    type_compatibility %>%
    filter(!compatible) %>%
    select(package_name, function_name) %>%
    distinct()

# Remove them.
type_analysis_data <-
    type_analysis_data %>%
    dplyr::filter(!(paste0(package_name, c("::"), function_name) %in%
                    paste0(incompatible_functions$package_name,
                           "::",
                           incompatible_functions$function_name)))

static_occ_by_type <- type_analysis_data %>% group_by(type) %>% 
  summarize(count = sum(count)) %>% 
  mutate(perc_arg = round(100 * count / sum(count), 2)) %>%
  arrange(-count)

# All ContractR assertions.
big_simplified_assertion_df <- read_csv("/mnt/nvme0/R/project-typeR/artifact/typeR/data/assertions-all.csv")
big_simplified_assertion_df %>% group_by(package_name, function_name, parameter_position, actual_type, expected_type) %>% summarize(count = n(), succ_ass = sum(assertion_status)) -> big_simplified_assertion_df_extra_simpl
big_simplified_assertion_df_extra_simpl <- read_csv(file.path(params$data_dir, "assertions-all-counts.csv"))

# The output of running propagatr, reduced the data (processed), and combined it into mega CSV.
core_corpus_run <- read_csv(file.path(params$data_dir, "assertions-all.csv"), col_types=paste0("cccccl", paste(rep("ccc", 21), collapse=""), "i")) %>% filter(has_dots != "has_dots")
# UNCOMMENT when paths are fixed
# core_corpus_run <- read_csv(file.path(base_path, "all.csv"), col_types=paste0("cccccl", paste(rep("ccc", 21), collapse=""), "i")) %>% filter(has_dots != "has_dots")
core_corpus_counts <- quick_gather_types(core_corpus_run)

# For generating some numbers.
assertions_by_fun <- read_csv("/mnt/arraySSD/alexi/oopsla_analysis_data/total_assertions_by_pkg_and_fun.csv")
# UNCOMMENT when paths fixed
assertions_by_fun <- read_csv(file.path(base_path, "total_assertions_by_pkg_and_fun.csv"))

# For Table 4
df_failed_assertions <- read_csv("/mnt/arraySSD/alexi/oopsla_analysis_data/failed-assertions.csv")
# UNCOMMENT when paths are fixed
# df_failed_assertions <- read_csv(file.path(base_path, "failed-assertions.csv"))
df_failed_assertions_no_undef <- df_failed_assertions %>% filter(expected_type != "<undefined>")

args_with_failed_assertions <- function_arg_counts %>% semi_join(df_failed_assertions_no_undef)
```

## Corpus -- Table 1

This bit of code generates Table 1 as seen in the paper, which includes the most common types seen across R functions.

```{r}
# dyn_occ_by_type <- big_simplified_assertion_df %>% mutate(count = 1) %>% group_by(actual_type) %>% summarize(count = sum(count)) %>% mutate(perc_arg = round(100 * count / sum(count), 2)) %>% arrange(-count) %>% rename("type" = "actual_type")
dyn_occ_by_type <- big_simplified_assertion_df_extra_simpl %>% mutate(static_count = 1) %>% group_by(actual_type) %>% summarize(static_count = sum(static_count)) %>% mutate(perc_arg = round(100 * static_count / sum(static_count), 2)) %>% arrange(-static_count) %>% rename("type" = "actual_type")

dyn_occ_by_type_core_corpus <- core_corpus_counts %>% group_by(type) %>% summarize(count = sum(count)) %>%
  filter(substr(type, 1, 3) %in% c(substr(static_occ_by_type %>% head(10) %>% select(type) %>% unlist %>% unname, 1, 3), "NUL"))

dyn_occ_by_type_core_corpus$type %>% map(function(s) strsplit(s, "@", fixed=T)[[1]][1]) -> type_names
dyn_occ_by_type_core_corpus$type <- type_names %>% unlist

dyn_occ_for_write <- dyn_occ_by_type_core_corpus %>% group_by(type) %>% summarize(count = sum(count)) %>% ungroup %>% filter(type %in% c(static_occ_by_type %>% head(10) %>% select(type) %>% unlist %>% unname, "NULL"))

dyn_occ_for_write$type[dyn_occ_for_write$type == "NULL"] <- "null"

dyn_occ_for_write <- dyn_occ_for_write %>% mutate(perc_arg = round(100 * count / sum(dyn_occ_by_type_core_corpus$count), 2))

corpus_tbl <- inner_join(static_occ_by_type, dyn_occ_for_write, by="type") %>%
  rename("Type" = "type", "Args with Type" = "count.x", "% of Static Types" = "perc_arg.x", "Dynamic Observations" = "count.y", "% of Dynamic Types" = "perc_arg.y")

# For generating the LaTeX. Unneeded ATM.
# kable(corpus_tbl, "latex", booktabs=T)

# Display.
corpus_tbl
```

## 6.1 -- Figures 6

This code generates Figure 6, showcasing the size of union types in our corpus.

```{r}
type_analysis_data <-
    type_analysis_data %>%
    mutate(type = map_chr(type, simplify_types)) %>%
    mutate(kind = map_chr(type, infer_kind))

union_data <-
type_analysis_data %>%
group_by(package_name, function_name, parameter_position) %>%
mutate(alternative_count = length(paste0(outer_alternative, "-", inner_alternative))) %>%
ungroup()
# %>%
#filter(alternative_count != 1)


union_frequency <-
union_data %>%
distinct(package_name, function_name, parameter_position, alternative_count) %>%
group_by(alternative_count) %>%
summarize(parameter_count = n()) %>%
ungroup() %>%
arrange(desc(parameter_count)) %>%
mutate(percentage = round(parameter_count / sum(parameter_count), 4),
       cumulative_percentage = round(cumsum(parameter_count) / sum(parameter_count), 4))

# datatable(union_frequency)

total_parameter_count <- sum(union_frequency$parameter_count)

# print(total_parameter_count)

union_frequency_folded <-
    union_frequency %>%
    select(alternative_count, parameter_count, percentage) %>%
    filter(alternative_count > 3) %>%
    summarize(alternative_count = "> 3", parameter_count = sum(parameter_count)) %>%
    mutate(percentage = round(as.double(parameter_count) / total_parameter_count, 3)) %>%
    print()
    
union_frequency_folded <-
    union_frequency %>%
    select(alternative_count, parameter_count, percentage) %>%
    filter(alternative_count <= 3) %>%
    mutate(alternative_count = as.character(alternative_count)) %>%
    rbind(union_frequency_folded)
    
# datatable(union_frequency_folded)
    
(union_frequency_folded %>%
ggplot(aes(alternative_count, percentage)) +
geom_col() +
scale_y_continuous(sec.axis = sec_axis(~ . * total_parameter_count)) +
labs(x="Types", y="Positions") +
scale_x_discrete(limits=c("> 3", 3, 2, 1)) +
coord_flip() +
theme(legend.title = element_blank(), legend.position = "top")) # %>%
```

## 6.1 -- Table 3

This table breaks down the singleton type categories found in our types.

```{r}
type_analysis_data %>% group_by(package_name, function_name, parameter_position, kind) %>% summarize(inner_alternative = sum(inner_alternative)) %>% filter(inner_alternative == 0) -> type_analysis_data_mono_pos

# Display.
type_analysis_data_mono_pos %>%
select(kind) %>%
group_by(kind) %>%
summarize(kind_count = n()) %>%
ungroup() %>%
arrange(desc(kind_count)) %>%
mutate(kind_perc = round(100 * kind_count / sum(kind_count), 2)) %>%
mutate(kind_cumperc = round(100 * cumsum(kind_count) / sum(kind_count), 2))
```

## 6.1 -- Figure 7

This figure explores function-level polymorphism.

```{r}
is_polymorphic <- function(kind) {
    kind %in% c("...", "any", "list")
}

total_function_count <- nrow(distinct(type_analysis_data, package_name, function_name))

function_polymorphism <-
union_data %>%
distinct(package_name, function_name, parameter_position, alternative_count) %>%
mutate(polymorphic = ifelse(alternative_count > 1, TRUE, is_polymorphic(alternative_count))) %>%
group_by(package_name, function_name) %>%
summarize(polymorphic_position_count = as.integer(sum(polymorphic))) %>%
ungroup()

package_polymorphism <-
function_polymorphism %>%
group_by(package_name) %>%
summarize(monomorphic = all(polymorphic_position_count == 0)) %>%
ungroup()

function_polymorphic_freq <-
function_polymorphism %>%
group_by(polymorphic_position_count) %>%
summarize(function_count = n()) %>%
ungroup() %>%
arrange(polymorphic_position_count)


function_polymorphic_freq_more_than_three <-
function_polymorphic_freq %>%
filter(polymorphic_position_count > 3) %>%
summarize(polymorphic_position_count = "> 3", function_count = sum(function_count))

function_polymorphic_freq <-
function_polymorphic_freq %>%
filter(polymorphic_position_count <= 3) %>%
rbind(function_polymorphic_freq_more_than_three) %>%
mutate(function_percentage = function_count / total_function_count)


(function_polymorphic_freq %>%
ggplot(aes(polymorphic_position_count, function_percentage)) +
geom_col() +
scale_y_continuous(sec.axis = sec_axis(~ . * total_function_count)) +
scale_x_discrete(limits=c("> 3", 3, 2, 1, 0)) +
labs(x="Polymorphic Positions", y="Functions") +
coord_flip() +
theme(legend.title = element_blank(), legend.position = "top")) # %>%
# save_graph("function_polymorphism.pdf")
```

## 6.2 -- Table 4

This table shows the most frequently occuring contract assertion failures we encountered when running our package client tests.

```{r}
df_failed_assertions_no_undef %>% 
  mutate(count = 1) %>%
  group_by(actual_type, expected_type) %>%
  summarize(count=sum(count)) %>%
  ungroup %>% 
  arrange(-count) %>%
  mutate(perc = round(100 * count / sum(count), 10)) %>%
  mutate(cumperc = round(100 * cumsum(perc) / sum(perc), 2)) %>%
  mutate(perc = round(perc, 2)) -> failed_assertions_raw_by_type

# For the LaTeX:
# kable(head( failed_assertions_raw_by_type %>%
#              rename(c("Actual Type" = "actual_type", "Expected Type" = "expected_type", "# Failed Assertions" = "count", "% Total" = "perc", "Cumulative %" = "cumperc")), 10), "latex", booktabs=T)

# Display.
failed_assertions_raw_by_type %>% head(10)
```

## 6.2 -- Table 5

This table breaks down the assertion failures statically, by looking at how many unique argument types were violated.

```{r}
head(args_with_failed_assertions %>% inner_join(df_failed_assertions_no_undef) %>% mutate(count = 1) %>% group_by(actual_type, expected_type) %>% summarize(count = sum(count)) %>% ungroup %>% arrange(-count), 10) %>% inner_join(args_with_failed_assertions %>% inner_join(df_failed_assertions_no_undef) %>% group_by(package_name, function_name, parameter_position, actual_type, expected_type) %>% summarize %>% ungroup %>% mutate(count = 1) %>% group_by(actual_type, expected_type) %>% summarize(count = sum(count)) %>% ungroup %>% arrange(-count), by=c("actual_type", "expected_type")) -> tbl_5_paper

sigs_to_look_for <- tbl_5_paper$expected_type

# arg_type_occurrence <- big_simplified_assertion_df %>% group_by(package_name, function_name, parameter_position, expected_type) %>% summarize() %>% ungroup %>% mutate(count = 1) %>% group_by(expected_type) %>% summarize(count = sum(count))
arg_type_occurrence <- big_simplified_assertion_df_extra_simpl %>% group_by(package_name, function_name, parameter_position, expected_type) %>% summarize() %>% ungroup %>% mutate(count = 1) %>% group_by(expected_type) %>% summarize(count = sum(count))

cross_ref <- arg_type_occurrence %>% filter(expected_type %in% sigs_to_look_for)

# Making Table 4
tbl_5_paper %>% inner_join(cross_ref) %>% select(actual_type, expected_type, count.y, count) %>% mutate(perc = round(100 * count.y / count, 2)) -> next_table

next_table
```

## 6.2 -- Tables 6 and 7

These next two figures look at the most frequently violated argument types.

```{r}
# arg_type_occurrence <- big_simplified_assertion_df %>% group_by(package_name, function_name, parameter_position, expected_type) %>% summarize() %>% ungroup %>% mutate(count = 1) %>% group_by(expected_type) %>% summarize(count = sum(count))
arg_type_occurrence <- big_simplified_assertion_df_extra_simpl %>% group_by(package_name, function_name, parameter_position, expected_type) %>% summarize() %>% ungroup %>% mutate(count = 1) %>% group_by(expected_type) %>% summarize(count = sum(count))

df_failed_assertions_no_undef %>% group_by(package_name, function_name, parameter_position, actual_type, expected_type) %>% summarize %>% ungroup %>% mutate(count = 1) %>% group_by(actual_type, expected_type) %>% summarize(count = sum(count)) %>% ungroup %>% arrange(-count) %>% inner_join(arg_type_occurrence, by="expected_type") %>% rename("sig_failed_count" = "count.x", "tot_sigs" = "count.y") -> arg_sig_viol_breakdown

arg_sig_viol_breakdown %>% filter(tot_sigs >= quantile(arg_sig_viol_breakdown$tot_sigs, .90)) %>%
  mutate(arg_type_failure_rate = sig_failed_count / tot_sigs) %>% arrange(-arg_type_failure_rate) %>% head(5) -> tbl6

tbl6
```

```{r}
arg_sig_viol_breakdown %>% filter(tot_sigs > quantile(arg_sig_viol_breakdown$tot_sigs, .82)) %>%
  mutate(arg_type_failure_rate = sig_failed_count / tot_sigs) %>% arrange(-arg_type_failure_rate) -> tbl7

tbl7 %>% head(5)
```

## Key Numbers

```{r}
perc_failed <- 1 - (big_simplified_assertion_df_extra_simpl %>% select(succ_ass) %>% sum) / (big_simplified_assertion_df_extra_simpl %>% select(count) %>% sum)

perc_failed
```

We have that `r perc_failed`% of assertions failed.

```{r}
prop_funs_with_a_failed_assertion <- nrow(df_failed_assertions_no_undef %>% group_by(package_name, function_name) %>% summarize() %>% ungroup) / nrow(assertions_by_fun)

prop_funs_with_a_failed_assertion
```

We have that `r prop_funs_with_a_failed_assertion`% of functions had their inferred contract violated.
